Decomposing biodiversity change to processes of extinction, colonization, and persistence across scales

### 2023-09-12 scripts/ms_scaling_changes.qmd
revision

```{r}
pdf("../figures/pCEP_2periods.pdf",
    height = 5, width = 5.5)
for(THE_metric in c("col", "ext")){
  print(predicted_pCEP_BBS %>% 
  mutate(spatial_scale = ifelse(AREA < 0.2, 0.13, 2.51)) %>%
  mutate(period = fct_relevel(period, c("M2M4", "M2M3", "M3M4"))) %>% 
  dplyr::select(spatial_scale, pred, period, metric) %>% 
  rbind(predicted_pCEP_atlas %>% dplyr::select(spatial_scale, pred, period, metric)) %>% 
  filter(pred >= 0 & pred <= 1) %>% 
  filter(period != "M2M4") %>%
  filter(metric == THE_metric) %>%
  ggplot()+
  geom_boxplot(aes(spatial_scale, pred, group = interaction(spatial_scale, period), color = period))+
  stat_summary(aes(spatial_scale, pred, color = period, linetype = period), fun.y = median, geom = 'line', linetype = 2, size = 1)+
  scale_x_log10()+
  annotation_logticks(sides = "b")+
  xlab("Spatial grain (km²)")+
  ylab(THE_metric)+
  ylim(0, .5)+
  labs(color = "Period", shape = "Period", linetype = "Period")+
  theme_bw()+
  theme(text = element_text(size = 16))+
  theme(panel.background = element_blank(),
        legend.position = "none",
        panel.grid.major = element_line(colour = "lightgrey"),
        axis.text.x = element_markdown()))
}
dev.off()
```

>>>>

```{r}
pdf("../figures/pCEP_2periods.pdf",
    height = 5, width = 5.5)
for(THE_metric in c("col", "ext")){
  print(predicted_pCEP_BBS %>% 
  mutate(spatial_scale = ifelse(AREA < 0.2, 0.13, 2.51)) %>%
  mutate(period = fct_relevel(period, c("M2M4", "M2M3", "M3M4"))) %>% 
  dplyr::select(spatial_scale, pred, period, metric) %>% 
  rbind(predicted_pCEP_atlas %>% dplyr::select(spatial_scale, pred, period, metric)) %>% 
  filter(pred >= 0 & pred <= 1) %>% 
  filter(period != "M2M4") %>%
  filter(metric == THE_metric) %>%
  ggplot()+
  geom_boxplot(aes(spatial_scale, pred, group = interaction(spatial_scale, period), color = period))+
  stat_summary(aes(spatial_scale, pred, color = period, linetype = period), fun.y = median, geom = 'line', linetype = 2, size = 1)+
  scale_x_log10()+
  annotation_logticks(sides = "b")+
  xlab("Spatial grain (km²)")+
  ylab(THE_metric)+
  ylim(0, .5)+
  labs(color = "Period", shape = "Period", linetype = "Period")+
  theme_bw()+
  theme(text = element_text(size = 16))+
  theme(panel.background = element_blank(),
        legend.position = "none",
        panel.grid.major = element_line(colour = "lightgrey"),
        axis.text.x = element_markdown()))
}

## Variable importance (Appendix B)

```{r}
CEP_atlas_cart_data <- readRDS("../data/data_model_fitting/CEP_atlas_cart_data.rds")
CEP_atlas_cart_data$metric[CEP_atlas_cart_data$metric == "proba_colonized"] <- "proba_col"

CEP_JPSP_cart_data <- readRDS("../data/data_model_fitting/CEP_JPSP_cart_data.rds")
CEP_JPSP_cart_data <- CEP_JPSP_cart_data %>% 
  filter(metric != "proba colonization") %>% 
  mutate(metric = case_when(
    metric == "colonization" ~ "col",
    metric == "extinction" ~ "ext",
    metric == "persistence" ~ "pers",
    metric == "proba persistence" ~ "proba_pers",
    metric == "proba extinction" ~ "proba_ext",
    metric == "proba colonized" ~ "proba_col",
  ))

pdf("../figures/correlations.pdf",
    height = 8.27, width = 11.69)
## For Atlas model
plot(atlas_cart_data[,names(atlas_cart_data) != "sr"],
     main = "Correlation - CzAtlas - species richness predictors")

ggplot(data = varImp(models_list_atlas$xgboost), aes(x = reorder(rownames(importance), Overall), y = Overall)) +
  geom_bar(stat = "identity", fill = "blue") +
  labs(x = "Predictor Variables", y = "Importance") +
  ggtitle("Variable Importance Plot - CzAtlas - XGBoost algorithm") +
  theme_bw()+
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

CEP_atlas_cart_data %>% filter(metric == "ext") %>% 
    select(-quadrat, -metric, -trials, -value) %>% 
    plot(main = "Correlation - CzAtlas - colonization/extinction/persistence")

for(THE_metric in unique(CEP_atlas_cart_data$metric)){
  if(!grepl("proba_", THE_metric)){
    print(
      ggplot(data = varImp(nCEP_atlas_models[[THE_metric]][["xgboost"]]), aes(x = reorder(rownames(importance), Overall), y = Overall)) +
      geom_bar(stat = "identity", fill = "blue") +
      labs(x = "Predictor Variables", y = "Importance") +
      ggtitle(paste0("Variable Importance Plot - CzAtlas - ", THE_metric)) +
      theme_bw()+
      theme(axis.text.x = element_text(angle = 45, hjust = 1))
    )
  }else{
    print(
      ggplot(data = varImp(pCEP_atlas_models[[gsub("proba_","",THE_metric)]][["xgboost"]]), aes(x = reorder(rownames(importance), Overall), y = Overall)) +
      geom_bar(stat = "identity", fill = "blue") +
      labs(x = "Predictor Variables", y = "Importance") +
      ggtitle(paste0("Variable Importance Plot - CzAtlas - ", THE_metric)) +
      theme_bw()+
      theme(axis.text.x = element_text(angle = 45, hjust = 1)) 
    )
  }
}

## For CzBBS model
plot(JPSP_cart_data[,names(JPSP_cart_data) != "sr"],
     main = "Correlation - CzBBS - species richness predictors")

ggplot(data = varImp(models_list_JPSP$rf), aes(x = reorder(rownames(importance), Overall), y = Overall)) +
  geom_bar(stat = "identity", fill = "blue") +
  labs(x = "Predictor Variables", y = "Importance") +
  ggtitle("Variable Importance Plot - CzBBS Data - random forest algorithm") +
  theme_bw()+
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

CEP_JPSP_cart_data %>% 
  filter(metric == "ext") %>%
  dplyr::select(-metric) %>% 
  filter(!is.na(value)) %>% 
  plot(main = paste0("Correlation - CzBBS - colonization/extinction/persistence predictors"))

for(THE_metric in unique(CEP_JPSP_cart_data$metric)){
  if(!grepl("proba_", THE_metric)){
    print(
      ggplot(data = varImp(nCEP_BBS_models[[THE_metric]][["rf"]]), aes(x = reorder(rownames(importance), Overall), y = Overall)) +
      geom_bar(stat = "identity", fill = "blue") +
      labs(x = "Predictor Variables", y = "Importance") +
      ggtitle(paste0("Variable Importance Plot - CzBBS - ", THE_metric)) +
      theme_bw()+
      theme(axis.text.x = element_text(angle = 45, hjust = 1))
    )
  }else{
    print(
      ggplot(data = varImp(pCEP_atlas_models[[gsub("proba_","",THE_metric)]][["xgboost"]]), aes(x = reorder(rownames(importance), Overall), y = Overall)) +
      geom_bar(stat = "identity", fill = "blue") +
      labs(x = "Predictor Variables", y = "Importance") +
      ggtitle(paste0("Variable Importance Plot - CzBBS - ", THE_metric)) +
      theme_bw()+
      theme(axis.text.x = element_text(angle = 45, hjust = 1)) 
    )
  }
}

dev.off()
```
